Summer 2025 ToDo

### 2025-07-21 Relations/Paper/Well-foundedness.tex
Updated typed lambda calculus code, also New outline of Section 5 in the ARS paper.

@@

>>>>

\newcommand{\accCor}{\textbf{accCor}}
\newcommand{\accDNE}{\textbf{accDNE}}

@@

\subsection{New version: Synopsis and Outline}

Classically, well-foundedness can be defined in several equivalent but distinct ways:
by induction, by absence of decreasing sequences, and by the fact that every
subset has a minimal element.  Constructively, these are no longer equivalent.
Moreover, even individual classical definitions can be given constructive
interpretations in varying degrees of logical strength.

For example, the implication from ``every non-empty subset has a minimal element'' to
``every element is accessible"  is a classical proof by contradiction.
(Assuming that there exists an inaccessible element, the hypothesis implies that
there must be a minimal such element, which immediately yields a contradiction:
every minimal element is accessible.)

Constructively, this argument only goes so far as to show that, if every non-empty
subset has a minimal element, then no element is inaccessible.
That is, the universal statement being proved is that every $x \in A$
is not-not-accessible. The final step needed to infer that $x$ is indeed accessible
is not constructively provable in general, therefore it needs to be assumed as
an additional hypothesis in order to recover the full implication.

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in many cases of interests.  In contrast, the original hypothesis
is generally equivalent to every subset being decidable.
(The file WFCounters.agda shows that assuming the full minimality hypothesis
for the standard strict order on the natural numbers entails decidability
of every subset of $\nat$, clearly incompatible with effective semantics.)

One could also consider proving well-foundedness ``up to $\lnot\lnot$",
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
This cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
On the other hand, the implication from the sequential to inductive definition
still requires an additional hypothesis, which may be described as the
constructive contrapositive of the definition of accessibility.
Specifically, we call {\accCor} the statement that, if an element $x$
is \emph{not} accessible, then there (strongly) exists a $y$ with $Ryx$
and $y$ not accessible.  Assuming \accCor, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

As this discussion illustrates, the network of relationships between the
constructive notions of well-foundedness is quite rich.  The purpose of
this section is to present the basic picture of these implications,
and elaborate on some of them.  Our key observations are summarized
in the following list.
\begin{enumerate}
  \item The implication from the inductive to the sequential definition
  requires the underlying relation to be decidable. This assumption is no longer
  necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires Markov's Principle
  (and decidability of $R$) to yield the same consequence.
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent,
  and imply WFacc- via WFseq-, requiring {\accCor} assumption in the final inference.
  \item At the same time, if the relation is finitely branching, then WFminDNE-
  implies WFacc- directly.
  \item Finite branchingness together with (weak) decidability of minimality
  implies \accCor.
  % \apcomment{The previous observation suggests the question of whether isWFseq- might
  % imply isWFminDNE- via decidability of minimal elements.\\
  % However, in trying to prove this implication, the hypothesis asserting non-existence of
  %  a minimal element of some given non-empty $\lnot\lnot$-closed predicate
  %  automatically entails that any particular element of interest is not normal.\\
  %  Unless it's not satisfying $P$? \emph{We need to decide whether it is P-minimal as well.}}
  \item Assuming \accCor, isWFacc- can also be derived from a weaker condition, isWFminCor(-).
  (Just like isWFminDNE is a natural weakening of isWFmin sufficient to recover isWFacc
  when {\accDNE} is available.)
  It is implied by isWFseq-, but does not appear to imply anything on its own.

  \apcomment{(This seems easy: The converse implication isWFminCor to isWFseq- requires the sequence in question to
  have decidable image, leading to a further weakening of isWFseq-.)}
  \item Since isWFminDNE- implies isWFseq-, the former also implies isWFminCor-.
  But the converse fails: If a subset/predicate $P$ is $\lnot\lnot$-closed,
  we cannot conclude it is coreductive.

  \item In general, the coreductive properties speak to the ability to choose
  a successor (strong existence).  This ability is central to being able to
  define sequences.  Therefore, sequential definitions generally require
  some coreductive situation to be applicable.

  \item Assuming normal forms are decidable,
  isWFseq implies a very weak version of minimality, isWFminEM.
  On its own this property is insufficient to imply anything else,
  so we have nothing else to say about it.
\end{enumerate}
